---
title: "Attention towards the different types of ambient traffic"
author: "Yee Thung Lee"
date: "2026-04-24"
output: html_document
---

# LOAD LIBRARIES
```{r setup, include=FALSE}
# LOAD LIBRARIES
suppressPackageStartupMessages({
  library(tidyr)
  library(dplyr)
  library (ggplot2)
  library (lmerTest)
  library (emmeans)
  library (lme4)
  library (viridis)
  library (readxl)
})
```

# READ EXCEL SHEET
```{r}
# LOAD EXCEL SHEET
# SUMMARY OF GLANCE DATA
SummaryOfGlanceData <- read_excel(
  "C:\\Users\\tsytl\\Desktop\\SUMMARY OF GLANCE.xlsx",
  sheet = "Sheet1"
)

head (SummaryOfGlanceData)
```

## DEFINE VARIABLE FOR DATA ANALYSIS
```{r}
# RENAME COLUMNS
SummaryOfGlanceData <- SummaryOfGlanceData %>%
  rename(
    LightingConditions = `Lighting Conditions`,
    AutomationLevel = `Automation Level`,
    SubNum = `Sub Number`,
    RunNum = `Run Number`,
    ParticipantID = `Participant Name`
  )

# RENAME COLUMNS FOR THE NUMBER OF GLANCE DATA
## I NEED THE FOLLOWING DATA FOR FURTHER ANALYSIS
### BEHAVIOURAL PREDICTION (BP) ELEMENT - WALKING PEDESTRIAN AND CYCLIST
### ENVIRONMENTAL PREDICTION (EP) ELEMENT - PARKED CARS AND PARKED TRUCKS
### LEFT - LEFT
### RIGHT - RIGHT
### 8 COLUMNS RENAME

SummaryOfGlanceData <- SummaryOfGlanceData %>%
  rename(
    NoOfGlance_LeftAmbientPedestrian_LeftPeriphery = `No Of Glances - Left Ambient Pedestrian - Left Periphery`,
    NoOfGlance_LeftCyclist_LeftPeriphery = `No Of Glances - Left Cyclist - Left Periphery`,
    NoOfGlance_LeftParkedCar_LeftPeriphery = `No Of Glances - Left Parked Car - Left Periphery`,
    NoOfGlance_LeftParkedTruck_LeftPeriphery = `No Of Glances - Left Parked Truck - Left Periphery`,
    NoOfGlance_RightAmbientPedestrian_RightPeriphery = `No Of Glances - Right Ambient Pedestrian - Right Periphery`,
    NoOfGlance_RightCyclist_RightPeriphery = `No Of Glances - Right Cyclist - Right Periphery`,
    NoOfGlance_RightParkedCar_RightPeriphery = `No Of Glances - Right Parked Car - Right Periphery`,
    NoOfGlance_RightParkedTruck_RightPeriphery = `No Of Glances - Right Parked Truck - Right Periphery`,
  )

# RENAME COLUMNS FOR THE MEAN GLANCE DURATION DATA
SummaryOfGlanceData <- SummaryOfGlanceData %>%
  rename(
    MGD_LeftAmbientPedestrian_LeftPeriphery = `MGD - Left Ambient Pedestrian - Left Periphery`,
    MGD_LeftCyclist_LeftPeriphery = `MGD - Left Cyclist - Left Periphery`,
    MGD_LeftParkedCar_LeftPeriphery = `MGD - Left Parked Car - Left Periphery`,
    MGD_LeftParkedTruck_LeftPeriphery = `MGD - Left Parked Truck - Left Periphery`,
    MGD_RightAmbientPedestrian_RightPeriphery = `MGD - Right Ambient Pedestrian - Right Periphery`,
    MGD_RightCyclist_RightPeriphery = `MGD - Right Cyclist - Right Periphery`,
    MGD_RightParkedCar_RightPeriphery = `MGD - Right Parked Car - Right Periphery`,
    MGD_RightParkedTruck_RightPeriphery = `MGD - Right Parked Truck - Right Periphery`,
  )
```

## FILTER THE DATASET TO ONLY HAVE THIS COLUMNS INCLUDED
```{r}
SummaryOfGlanceDataClean <- SummaryOfGlanceData %>%
  select(
    LightingConditions,
    AutomationLevel,
    SubNum,
    RunNum,
    ParticipantID,
    
    # Glance counts
    NoOfGlance_LeftAmbientPedestrian_LeftPeriphery,
    NoOfGlance_LeftCyclist_LeftPeriphery,
    NoOfGlance_LeftParkedCar_LeftPeriphery,
    NoOfGlance_LeftParkedTruck_LeftPeriphery,
    NoOfGlance_RightAmbientPedestrian_RightPeriphery,
    NoOfGlance_RightCyclist_RightPeriphery,
    NoOfGlance_RightParkedCar_RightPeriphery,
    NoOfGlance_RightParkedTruck_RightPeriphery,
    
    # Mean glance duration
    MGD_LeftAmbientPedestrian_LeftPeriphery,
    MGD_LeftCyclist_LeftPeriphery,
    MGD_LeftParkedCar_LeftPeriphery,
    MGD_LeftParkedTruck_LeftPeriphery,
    MGD_RightAmbientPedestrian_RightPeriphery,
    MGD_RightCyclist_RightPeriphery,
    MGD_RightParkedCar_RightPeriphery,
    MGD_RightParkedTruck_RightPeriphery
  )

```

## REORGANISE DATA FOR ANALYSIS
```{r}
SummaryOfGlanceDataCleanForDataAnalysis <- SummaryOfGlanceDataClean %>%
  pivot_longer(
    cols = starts_with("NoOfGlance") | starts_with("MGD"),
    names_to = c("Measure", "Condition"),
    names_pattern = "^(NoOfGlance|MGD)_(.*)$",
    values_to = "Value"
  )


SummaryOfGlanceDataCleanForDataAnalysis <- SummaryOfGlanceDataCleanForDataAnalysis %>%
  mutate(
    Side = ifelse(grepl("Left", Condition), "Left", "Right"),
    Element = case_when(
      grepl("Cyclist", Condition) ~ "Cyclist",
      grepl("Pedestrian", Condition) ~ "Pedestrian",
      grepl("ParkedCar", Condition) ~ "ParkedCar",
      grepl("ParkedTruck", Condition) ~ "ParkedTruck"
    ),
    ElementType = case_when(
      Element %in% c("Cyclist", "Pedestrian") ~ "Behavioural Prediction (BP)",
      Element %in% c("ParkedCar", "ParkedTruck") ~ "Inanimate Object (IO)"
    )
  )


```


### REORGANISE THE DATA

```{r cars}
bp_ep_data <- SummaryOfGlanceDataCleanForDataAnalysis %>%
  group_by(ParticipantID, SubNum, RunNum, LightingConditions, AutomationLevel, Side, ElementType, Measure) %>%
  summarise(
    Value = mean(Value, na.rm = TRUE),
    .groups = "drop"
  )

bp_ep_glance <- bp_ep_data %>%
  filter(Measure == "NoOfGlance")

bp_ep_mgd <- bp_ep_data %>%
  filter(Measure == "MGD")
```

### EXPORT DATA

```{r pressure, echo=FALSE}
## MEAN GLANCE DURATION
## FILTER DATA TO LEFT ONLY
left_MGD_data <- bp_ep_mgd %>%
  filter(Side == "Left")

## FILTER DATA TO RIGHT ONLY
right_MGD_data <- bp_ep_mgd %>%
  filter(Side == "Right")
```



```{r}
# FINAL MODEL
# LIGHT, AUTOMATION, ELEMENT TYPE, LIGHT*AUTOMATION, ELEMENTTYPE*LIGHTING, ELEMENTTYPE*AUTO

glmm_final_model_gamma_MeanGlanceDurationLEFTLEFT <- glmer(Value~ LightingConditions + AutomationLevel  + ElementType + LightingConditions*AutomationLevel + LightingConditions*ElementType + AutomationLevel*ElementType + (1|SubNum), family = Gamma (link = "log"), data = left_MGD_data)

summary (glmm_final_model_gamma_MeanGlanceDurationLEFTLEFT)
```

```{r}
library (performance)
# CALCULATE R-SQUARED VALUES 
r2_values <- r2(glmm_final_model_gamma_MeanGlanceDurationLEFTLEFT)
print (r2_values)
```


### Post-Hoc Comparison

#### Lighting of the Driving Environment

```{r}

# OBTAIN ESTIMATED MARGINAL MEANS (LIGHTING) - PAIRWISE 
lighting_emmeans <- emmeans (glmm_final_model_gamma_MeanGlanceDurationLEFTLEFT, ~ LightingConditions, type = "response")
lighting_emmeans
# PAIRWISE
pairwise_lighting <- pairs (lighting_emmeans, adjust = "bonferroni", type = "response")
pairwise_lighting

```

#### Level of Automation

```{r}
# OBTAIN ESTIMATED MARGINAL MEANS (AUTOMATION) - PAIRWISE 
automation_emmeans <- emmeans (glmm_final_model_gamma_MeanGlanceDurationLEFTLEFT, ~ AutomationLevel, type = "response")
automation_emmeans
# PAIRWISE
pairwise_automation <- pairs (automation_emmeans, adjust = "bonferroni", type = "response")
pairwise_automation
```

#### Element Type 
```{r}
# OBTAIN ESTIMATED MARGINAL MEANS (AUTOMATION) - PAIRWISE 
elementtype_emmeans <- emmeans (glmm_final_model_gamma_MeanGlanceDurationLEFTLEFT, ~ ElementType, type = "response")
elementtype_emmeans
# PAIRWISE
pairwise_elementtype <- pairs (elementtype_emmeans, adjust = "bonferroni", type = "response")
pairwise_elementtype
```

#### Plots
```{r}
library(emmeans)

emm1 <- emmeans(
  glmm_final_model_gamma_MeanGlanceDurationLEFTLEFT,
  ~ LightingConditions * AutomationLevel * ElementType
)

emm_df_1 <- as.data.frame(
  summary(emm1, type = "response")
)

emm_df_1



```


```{r}
library(ggplot2)
library(dplyr)

# Create ordered grouping variable
emm_df_1 <- emm_df_1 %>%
  mutate(Group = factor(
    paste(AutomationLevel, LightingConditions),
    levels = c(
      "Manual Driving Daytime",
      "Manual Driving AfterDark",
      "L2 Automation Daytime",
      "L2 Automation AfterDark"
    )
  ))

# Plot
plot_1 <- ggplot(emm_df_1, aes(x = Group, y = response, fill = ElementType)) +
  
  # bars
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  
  # error bars
  geom_errorbar(
    aes(ymin = asymp.LCL, ymax = asymp.UCL),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  
  # color mapping
  scale_fill_manual(values = c(
    "Behavioural Prediction (BP)" = "#D55E00", # Vermillion (Warm/Active)
    "Environmental Prediction (EP)" = "#56B4E9"  # Sky Blue (Cool/Static)
  )) +
  scale_y_continuous(breaks = c(0, 250, 500, 750, 1000)) +
  labs(x = "", 
    y = "Predicted Mean Glance Duration \n for Left Periphery (milliseconds)",
    fill = "Element Type"
  ) +
  theme_minimal() +
  theme(legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(face = "bold", size = 10),
    axis.title.y = element_text(face = "bold", colour = "black", size = 12),
    axis.text.y = element_text(face = "bold", color = "black", size = 12),
    panel.grid = element_blank(),    # remove grid
    panel.background = element_blank(), # remove panel background
    plot.background = element_blank(),  # remove outer background
    axis.ticks.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
    axis.line = element_line(color = "black"))

plot_1
```


```{r}
# FINAL MODEL
# LIGHT, AUTOMATION, ELEMENT TYPE, LIGHT*AUTOMATION, ELEMENTTYPE*LIGHTING, ELEMENTTYPE*AUTO

glmm_final_model_gamma_MeanGlanceDurationRIGHTRIGHT <- glmer(Value~ LightingConditions + AutomationLevel  + ElementType + LightingConditions*AutomationLevel + LightingConditions*ElementType + AutomationLevel*ElementType + (1|SubNum), family = Gamma (link = "log"), data = right_MGD_data)

summary (glmm_final_model_gamma_MeanGlanceDurationRIGHTRIGHT)
```


```{r}
library (performance)
# CALCULATE R-SQUARED VALUES 
r2_values <- r2(glmm_final_model_gamma_MeanGlanceDurationRIGHTRIGHT)
print (r2_values)
```


### Post-Hoc Comparison

#### Lighting of the Driving Environment

```{r}

# OBTAIN ESTIMATED MARGINAL MEANS (LIGHTING) - PAIRWISE 
lighting_emmeans <- emmeans (glmm_final_model_gamma_MeanGlanceDurationRIGHTRIGHT, ~ LightingConditions, type = "response")
lighting_emmeans
# PAIRWISE
pairwise_lighting <- pairs (lighting_emmeans, adjust = "bonferroni", type = "response")
pairwise_lighting

```


#### Level of Automation

```{r}
# OBTAIN ESTIMATED MARGINAL MEANS (AUTOMATION) - PAIRWISE 
automation_emmeans <- emmeans (glmm_final_model_gamma_MeanGlanceDurationRIGHTRIGHT, ~ AutomationLevel, type = "response")
automation_emmeans
# PAIRWISE
pairwise_automation <- pairs (automation_emmeans, adjust = "bonferroni", type = "response")
pairwise_automation
```



#### Element Type 
```{r}
# OBTAIN ESTIMATED MARGINAL MEANS (AUTOMATION) - PAIRWISE 
elementtype_emmeans <- emmeans (glmm_final_model_gamma_MeanGlanceDurationRIGHTRIGHT, ~ ElementType, type = "response")
elementtype_emmeans
# PAIRWISE
pairwise_elementtype <- pairs (elementtype_emmeans, adjust = "bonferroni", type = "response")
pairwise_elementtype
```

#### Interaction Effect (Element\*Lighting)

```{r}
# OBTAIN ESTIMATED MARGINAL MEANS (AUTOMATION*LIGHTTING) - PAIRWISE 
interaction_element_lighting_emmeans <- emmeans (glmm_final_model_gamma_MeanGlanceDurationRIGHTRIGHT, ~ ElementType*LightingConditions, type = "response")
interaction_element_lighting_emmeans

# PAIRWISE
pairwise_interaction <- pairs (interaction_element_lighting_emmeans, adjust = "bonferroni", type = "response")
pairwise_interaction
```


#### Plots
```{r}
library(emmeans)

emm2 <- emmeans(
  glmm_final_model_gamma_MeanGlanceDurationRIGHTRIGHT,
  ~ LightingConditions * AutomationLevel * ElementType
)

emm_df_2 <- as.data.frame(
  summary(emm2, type = "response")
)

emm_df_2


```

```{r}
library(ggplot2)
library(dplyr)

# Create ordered grouping variable
emm_df_2 <- emm_df_2 %>%
  mutate(Group = factor(
    paste(AutomationLevel, LightingConditions),
    levels = c(
      "Manual Driving Daytime",
      "Manual Driving AfterDark",
      "L2 Automation Daytime",
      "L2 Automation AfterDark"
    )
  ))

# Plot
plot_2 <- ggplot(emm_df_2, aes(x = Group, y = response, fill = ElementType)) +
  
  # bars
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  
  # error bars
  geom_errorbar(
    aes(ymin = asymp.LCL, ymax = asymp.UCL),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  
  # color mapping
  scale_fill_manual(values = c(
    "Behavioural Prediction (BP)" = "#D55E00", # Vermillion (Warm/Active)
    "Environmental Prediction (EP)" = "#56B4E9"  # Sky Blue (Cool/Static)
  )) +
  ylim (0, 1000) + 
  labs(x = "", 
    y = "Predicted Mean Glance Duration \n for Right Periphery (milliseconds)",
    fill = "Element Type"
  ) +
  theme_minimal() +
  theme(legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(face = "bold", size = 10),
    axis.title.y = element_text(face = "bold", colour = "black", size = 12),
    axis.text.y = element_text(face = "bold", color = "black", size = 12),
    panel.grid = element_blank(),    # remove grid
    panel.background = element_blank(), # remove panel background
    plot.background = element_blank(),  # remove outer background
    axis.ticks.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
    axis.line = element_line(color = "black"))

plot_2
```


