---
title: "Percentage of Glance Time (Forward Roadway) (%)"
author: "Yee Thung Lee"
date: "2025-11-30"
output: html_document
---

## Load Libraries

```{r setup, include=FALSE}
# LOAD LIBRARIES
suppressPackageStartupMessages({
  library(tidyr)
  library(dplyr)
  library (ggplot2)
  library (lmerTest)
  library (emmeans)
  library (lme4)
  library (viridis)
  library (readxl)
})
```

```{r}
# LOAD EXCEL SHEET
# % OF GLANCE TIME
PercentageOfGlanceTimeData <- read_excel(
  "C:\\Users\\tsytl\\OneDrive - University of Leeds\\Paper 2 - Visual Scanning Patterns\\PERCENTAGE OF GLANCE TIME ON EACH AOI.xlsx",
  sheet = "Sheet1"
)

head (PercentageOfGlanceTimeData)
```

## Define Variables for Data Analysis

```{r}
# DEFINE VARIABLE FOR DATA ANALYSIS

# RENAME COLUMNS
PercentageOfGlanceTimeData <- PercentageOfGlanceTimeData %>%
  rename(
    LightingConditions = `Lighting Conditions`,
    AutomationLevel = `Automation Level`,
    SubNum = `Sub Number`,
    RunNum = `Run Number`,
    PercentageOfGlanceOnRoad = `Percentage of Glance Time (On Road) (%)`
  )

# DEFINE COLUMNS AS FACTOR (SUBJECT ID, RUN NUMBER,INDEPENDENT VARIABLES)
PercentageOfGlanceTimeData$SubNum <- factor (PercentageOfGlanceTimeData$SubNum)
PercentageOfGlanceTimeData$RunNum <- factor (PercentageOfGlanceTimeData$RunNum)
PercentageOfGlanceTimeData$LightingConditions <- factor (PercentageOfGlanceTimeData$LightingConditions)
PercentageOfGlanceTimeData$AutomationLevel <- factor (PercentageOfGlanceTimeData$AutomationLevel)


# DEFINE COLUMNS AS DEPENDENT VARIABLES (% GLANCE TIME ON ROAD)
str(PercentageOfGlanceTimeData$PercentageOfGlanceOnRoad)
PercentageOfGlanceTimeData$PercentageOfGlanceOnRoad<- as.numeric(PercentageOfGlanceTimeData$PercentageOfGlanceOnRoad) 
```


## Final Model (REML = TRUE)

```{r}
# FINAL MODEL
# LIGHT, AUTOMATION, LIGHT*AUTOMATION

lmm_final_model_PercentageOfGlanceOnRoad <- lmer(PercentageOfGlanceOnRoad~ LightingConditions + AutomationLevel + LightingConditions*AutomationLevel  + (1|SubNum), data = PercentageOfGlanceTimeData)

summary (lmm_final_model_PercentageOfGlanceOnRoad)
anova (lmm_final_model_PercentageOfGlanceOnRoad)
```
### R2 Values

- The marginal R² represents the proportion of variance explained only by the fixed effects in your model.
- The conditional R² represents the proportion of variance explained by both the fixed effects and the random effects combined.

```{r}
library (performance)
# CALCULATE R-SQUARED VALUES 
r2_values <- r2(lmm_final_model_PercentageOfGlanceOnRoad)
print (r2_values)
```

## Assumption Check

```{r}
# ASSUMPTION CHECK
# LIBRARY FOR NORMALITY TEST
library (stats)

residual_lmm_final_model_PercentageOfGlanceOnRoad <- resid(lmm_final_model_PercentageOfGlanceOnRoad)

# KOMOLGOROV-SMIRNOV TEST (IF IS SIGNIFICANT, THEN IS NOT NORMALLY DISTRIBUTED - ASSUMPTION VIOLATED)
ks_test <- ks.test (residual_lmm_final_model_PercentageOfGlanceOnRoad, "pnorm", mean (residual_lmm_final_model_PercentageOfGlanceOnRoad), sd (residual_lmm_final_model_PercentageOfGlanceOnRoad))
print (ks_test)


# QQ PLOT (DEVIATION FROM THE DIAGONAL LINE INDICATE NOT NORMALLY DISTRIBUTED - ASSUMPTION VIOLATED)
qqnorm(residual_lmm_final_model_PercentageOfGlanceOnRoad)
qqline(residual_lmm_final_model_PercentageOfGlanceOnRoad, col = "red")

# Residuals vs Fitted
plot(fitted(lmm_final_model_PercentageOfGlanceOnRoad),
     resid(lmm_final_model_PercentageOfGlanceOnRoad),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted Values")
abline(h = 0, col = "red")
```

#### Notes - Assumption Met

### Post-Hoc Comparison

#### Lighting of the Driving Environment

```{r}

# OBTAIN ESTIMATED MARGINAL MEANS (LIGHTING) - PAIRWISE 
lighting_emmeans <- emmeans (lmm_final_model_PercentageOfGlanceOnRoad, ~ LightingConditions)
lighting_emmeans
# PAIRWISE
pairwise_lighting <- pairs (lighting_emmeans, adjust = "bonferroni")
pairwise_lighting

```

#### Level of Automation

```{r}
# OBTAIN ESTIMATED MARGINAL MEANS (AUTOMATION) - PAIRWISE 
automation_emmeans <- emmeans (lmm_final_model_PercentageOfGlanceOnRoad, ~ AutomationLevel, type = "response")
automation_emmeans
# PAIRWISE
pairwise_automation <- pairs (automation_emmeans, adjust = "bonferroni", type = "response")
pairwise_automation
```

##### Explanation

-   L2 automation reduced on-road glance time significantly by 3.81 % compared to manual driving (L2: M = 48.8%, SE = 1.32; Manual: M = 52.6%, SE = 1.29).

#### Interaction Effect (Automation\*Lighting)

```{r}
# OBTAIN ESTIMATED MARGINAL MEANS (AUTOMATION*LIGHTTING) - PAIRWISE 
interaction_automation_lighting_emmeans <- emmeans (lmm_final_model_PercentageOfGlanceOnRoad, ~ AutomationLevel*LightingConditions)
interaction_automation_lighting_emmeans

# PAIRWISE
pairwise_interaction <- pairs (interaction_automation_lighting_emmeans, adjust = "bonferroni")
pairwise_interaction
```






