---
title: "Percentage of Glance Time (Right Periphery) (%)"
author: "Yee Thung Lee"
date: "2025-11-30"
output: html_document
---

## Load Libraries

```{r setup, include=FALSE}
# LOAD LIBRARIES
suppressPackageStartupMessages({
  library(tidyr)
  library(dplyr)
  library (ggplot2)
  library (lmerTest)
  library (emmeans)
  library (lme4)
  library (viridis)
  library (readxl)
})
```

```{r}
# LOAD EXCEL SHEET
# % OF GLANCE TIME
PercentageOfGlanceTimeData <- read_excel(
  "C:\\Users\\tsytl\\OneDrive - University of Leeds\\Paper 2 - Visual Scanning Patterns\\PERCENTAGE OF GLANCE TIME ON EACH AOI.xlsx",
  sheet = "Sheet1"
)

head (PercentageOfGlanceTimeData)
```

## Define Variables for Data Analysis

```{r}
# DEFINE VARIABLE FOR DATA ANALYSIS

# RENAME COLUMNS
PercentageOfGlanceTimeData <- PercentageOfGlanceTimeData %>%
  rename(
    LightingConditions = `Lighting Conditions`,
    AutomationLevel = `Automation Level`,
    SubNum = `Sub Number`,
    RunNum = `Run Number`,
    PercentageOfGlanceRightPeriphery = `Percentage of Glance Time (Right Periphery) (%)`
  )

# DEFINE COLUMNS AS FACTOR (SUBJECT ID, RUN NUMBER,INDEPENDENT VARIABLES)
PercentageOfGlanceTimeData$SubNum <- factor (PercentageOfGlanceTimeData$SubNum)
PercentageOfGlanceTimeData$RunNum <- factor (PercentageOfGlanceTimeData$RunNum)
PercentageOfGlanceTimeData$LightingConditions <- factor (PercentageOfGlanceTimeData$LightingConditions)
PercentageOfGlanceTimeData$AutomationLevel <- factor (PercentageOfGlanceTimeData$AutomationLevel)


# DEFINE COLUMNS AS DEPENDENT VARIABLES (% GLANCE TIME ON ROAD)
str(PercentageOfGlanceTimeData$PercentageOfGlanceRightPeriphery)
PercentageOfGlanceTimeData$PercentageOfGlanceRightPeriphery<- as.numeric(PercentageOfGlanceTimeData$PercentageOfGlanceRightPeriphery) 
```

## Final Model (REML = TRUE)

```{r}
# FINAL MODEL
# LIGHT, AUTOMATION, RUN, LIGHT*AUTOMATION

lmm_final_model_PercentageOfGlanceRightPeriphery <- lmer(PercentageOfGlanceRightPeriphery~ LightingConditions + AutomationLevel + RunNum + LightingConditions*AutomationLevel + (1|SubNum), data = PercentageOfGlanceTimeData)

summary (lmm_final_model_PercentageOfGlanceRightPeriphery)
anova (lmm_final_model_PercentageOfGlanceRightPeriphery)
```
### R2 Values

- The marginal R² represents the proportion of variance explained only by the fixed effects in your model.
- The conditional R² represents the proportion of variance explained by both the fixed effects and the random effects combined.

```{r}
library (performance)
# CALCULATE R-SQUARED VALUES 
r2_values <- r2(lmm_final_model_PercentageOfGlanceRightPeriphery)
print (r2_values)
```

## Assumption Check

```{r}
# ASSUMPTION CHECK
# LIBRARY FOR NORMALITY TEST
library (stats)

residual_lmm_final_model_PercentageOfGlanceRightPeriphery <- resid(lmm_final_model_PercentageOfGlanceRightPeriphery)

# KOMOLGOROV-SMIRNOV TEST (IF IS SIGNIFICANT, THEN IS NOT NORMALLY DISTRIBUTED - ASSUMPTION VIOLATED)
ks_test <- ks.test (residual_lmm_final_model_PercentageOfGlanceRightPeriphery, "pnorm", mean (residual_lmm_final_model_PercentageOfGlanceRightPeriphery), sd (residual_lmm_final_model_PercentageOfGlanceRightPeriphery))
print (ks_test)


# QQ PLOT (DEVIATION FROM THE DIAGONAL LINE INDICATE NOT NORMALLY DISTRIBUTED - ASSUMPTION VIOLATED)
qqnorm(residual_lmm_final_model_PercentageOfGlanceRightPeriphery)
qqline(residual_lmm_final_model_PercentageOfGlanceRightPeriphery, col = "red")

# Residuals vs Fitted
plot(fitted(lmm_final_model_PercentageOfGlanceRightPeriphery),
     resid(lmm_final_model_PercentageOfGlanceRightPeriphery),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted Values")
abline(h = 0, col = "red")
```

#### Notes - Assumption Met

### Post-Hoc Comparison

#### Lighting of the Driving Environment

```{r}

# OBTAIN ESTIMATED MARGINAL MEANS (LIGHTING) - PAIRWISE 
lighting_emmeans <- emmeans (lmm_final_model_PercentageOfGlanceRightPeriphery, ~ LightingConditions)
lighting_emmeans
# PAIRWISE
pairwise_lighting <- pairs (lighting_emmeans, adjust = "bonferroni")
pairwise_lighting

```

#### Level of Automation

```{r}
# OBTAIN ESTIMATED MARGINAL MEANS (AUTOMATION) - PAIRWISE 
automation_emmeans <- emmeans (lmm_final_model_PercentageOfGlanceRightPeriphery, ~ AutomationLevel)
automation_emmeans
# PAIRWISE
pairwise_automation <- pairs (automation_emmeans, adjust = "bonferroni")
pairwise_automation
```

##### Explanation (Level of Automation)

-   L2 automation increased glance time towards right periphery significantly by 4.09 % compared to manual driving (L2 Automation: EMM = 20.0 %, SE = 0.77; Manual Driving: EMM = 15.9 %, SE = 0.73).

#### Run Number
```{r}
# OBTAIN ESTIMATED MARGINAL MEANS (RUN NUMBER) - PAIRWISE 
run_emmeans <- emmeans (lmm_final_model_PercentageOfGlanceRightPeriphery, ~ RunNum)
run_emmeans
# PAIRWISE
pairwise_run <- pairs (run_emmeans, adjust = "bonferroni")
pairwise_run
```
##### Explanation (Run Number)
- Drivers' percentage of glance time towards right periphery reduced significantly from Run 1 to Run 2 (by 5.80 %, t = 4.54, p < .001) and Run 3 (by 6.46 %, t = 5.08, p < .001).
- Drivers' percentage of glance time towards right periphery increased significantly from Run 2 to Run 4 (by 4.15 %, t = -3.18, p = .011), and from Run 3 to Run 4 (by 4.81 %, t = -3.71, p = .002)


#### Interaction Effect (Automation\*Lighting)

```{r}
# OBTAIN ESTIMATED MARGINAL MEANS (AUTOMATION*LIGHTTING) - PAIRWISE 
interaction_automation_lighting_emmeans <- emmeans (lmm_final_model_PercentageOfGlanceRightPeriphery, ~ AutomationLevel*LightingConditions)
interaction_automation_lighting_emmeans

# PAIRWISE
pairwise_interaction <- pairs (interaction_automation_lighting_emmeans, adjust = "bonferroni")
pairwise_interaction
```




